Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures

### 2017-03-27 qnml.tex
Typos Please enter the commit message for your changes. Lines starting

We will next state this as a
theorem and sketch a proof for it. A more detailed proof appears in the
supplementary material.
\begin{theorem}
  \label{thm:scoreqv}
  Let $G$ and $G'$ be two Bayesian network structures that differ only
  by a single covered arc reversal, i.e., the arc from $A$ to $B$ in $G$
  has been reversed in $G'$ to point from $B$ to $A$, then
  $$s^{qNML}(D;G)=s^{qNML}(D;G').$$
\end{theorem}
\begin{proof}
  Now the scores for structures can be decomposed as the
  $s^{qNML}(D;G)=\sum_{i=1}^{n}s_i^{qNML}(D;G)$ and
  $s^{qNML}(D;G')=\sum_{i=1}^{n}s_i^{qNML}(D;G')$.  Since only the
  terms corresponding to the variables $A$ and $B$ in these sums are
  different, it is enough to show that sum of these two terms are
  equal for $G$ and $G'$. Since we can assume the data to be fixed we
  lighten up the notations and write
  $P^1_{NML}(i,G_i) := P^1_{NML}(D[\cdot,(i,G_i)];G)$ and
  $P^1_{NML}(G_i)   := P^1_{NML}(D[\cdot,G_i];G)$.
  \begin{eqnarray}
    \lefteqn{s_A^{qNML}(D;G)+s_B^{qNML}(D;G)} \nonumber\\
    && =\log\frac{P^1_{NML}(A,G_{A})}{P^1_{NML}(G_{A})}
            \frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{B})}\nonumber\\
    && =\log 1\cdot\frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{A})}\nonumber\\
    && =\log \frac{P^1_{NML}(B,G'_{B})}{P^1_{NML}(G'_{A})}
             \frac{P^1_{NML}(A,G'_{A})}{P^1_{NML}(G'_{B})}\nonumber\\
 && =s_A^{qNML}(D;G')+s_B^{qNML}(D;G'),\nonumber
\end{eqnarray}
  due to equations
  $\{A\}\cup G_A  = G_B$,
  $\{B\}\cup G'_B = G'_A$,
  $\{B\}\cup G_B  = \{A\} \cup G'_A$,
  and $G_A        = G'_B$.
\end{proof}

@@

Clearly, the qNML score also decomposes by the connected components,
so it remains to show that if the (sub)network $G$ is a tournament,
then for any data $D$, $s^{qNML}(D;G)=\log P_{NML}(D;G)$.  Due to the
score equivalence of both the NML criterion and the qNML criterion, we
may pick such a tournament $G$ that linear ordering defined by $G$
matches the ordering of the data columns, i.e., $i<j$ implies $G_i
\subset G_j$. Now from the definition~(\ref{eqn:qnmldef}) of the qNML
criterion we see that for the tournament $G$, the sum telescopes
leaving us with $s^{qNML}(D;G) = \log P^1_{NML}(D[\cdot,G];G)$, thus
it is enough to show that $P^1_{NML}(D;G)=P_{NML}(D;G)$.
This follows, since for any data vector $x$ in data $D$, the
$P^1(x;\hat\theta(D),G) = P(x;\hat\theta(D),G)$,
where we have denoted by $P^1$ the model that takes $n$-dimensional
vectors to be values of the single (collapsed) categorical variable.
Denoting prefixes of data vector $x$ by $x^{:i}$, and number of times
such a prefix appears in a $N\times n$ datamatrix $D$ by $n_D(x^{:i})$,
($n_D(x^0)=N)$, we have
\begin{eqnarray}
  P(D;\hat\theta(D),G) &=& \prod_{j=1}^{N}P(d_j;\theta(D),G) \nonumber \\
  &=& \prod_{j=1}^{N} \prod_{i=1}^{n} \frac{n_D(d_j^{:i})}{n_D(d_j^{:i-1})} \nonumber\\
  &=& \prod_{j=1}^{N} \frac{n_D(d_j^{:n})}{N} \nonumber \\
  &=& P^1(D;\hat\theta^1(D),G).
\end{eqnarray}
Since both $P^1_{NML}$ and $P_{NML}$ are defined in terms of the
maximum likelihood probabilities, the equality above implies the equality
of these distributions.

@@

Collapsing many variables into a single one causes $r$ in the
equation~(\ref{eqn:szp}) to be exponentially large (in our experiments
around $2^{20}$).  This causes the $\Gamma^2$ terms to grow very large
calling for care when computing the ratios of such huge numbers. More
specifically, scoring the $i^\text{th}$ data column, i.e., computing
$s^{qNML}_i$, requires the difference $dreg(N,q_i,r_i)=reg(N,q_i) -
reg(N,r_i q_i)$, where $q$ denotes the number of values in the
collapsed parent variable.  A numerically stable approximation is:
\begin{eqnarray}
\lefteqn{dreg(N,q,r)\approx\frac{1-r^2}{18 N} q^{2}} \\
&&+ \frac{\left(q - 1\right) \sqrt{4 q - 2} + \left(1 - q r\right) \sqrt{4 q r - 2}}{6 \sqrt{N}} \nonumber\\
&&+\frac{q\left(3 N \left(r \log{r}  - \left(r - 1\right) \left(\log{N} + 1\right)\right) + r\right)}{6 N} \nonumber\\
&&+ \frac{1}{12}(\log{(r+\frac{1-r}{q+1})}- 7\log{r}).\nonumber
\end{eqnarray}

>>>>

We will next state this as a
theorem and sketch a proof for it. A more detailed proof appears in the
supplementary material.
\begin{theorem}
  \label{thm:scoreqv}
  Let $G$ and $G'$ be two Bayesian network structures that differ only
  by a single covered arc reversal, i.e., the arc from $A$ to $B$ in $G$
  has been reversed in $G'$ to point from $B$ to $A$, then
  $$s^{qNML}(D;G)=s^{qNML}(D;G').$$
\end{theorem}
\begin{proof}
  Now the scores for structures can be decomposed as the
  $s^{qNML}(D;G)=\sum_{i=1}^{n}s_i^{qNML}(D;G)$ and
  $s^{qNML}(D;G')=\sum_{i=1}^{n}s_i^{qNML}(D;G')$.  Since only the
  terms corresponding to the variables $A$ and $B$ in these sums are
  different, it is enough to show that sum of these two terms are
  equal for $G$ and $G'$. Since we can assume the data to be fixed we
  lighten up the notations and write
  $P^1_{NML}(i,G_i) := P^1_{NML}(D[\cdot,(i,G_i)];G)$ and
  $P^1_{NML}(G_i)   := P^1_{NML}(D[\cdot,G_i];G)$.
  \begin{eqnarray}
    \lefteqn{s_A^{qNML}(D;G)+s_B^{qNML}(D;G)} \nonumber\\
    && =\log\frac{P^1_{NML}(A,G_{A})}{P^1_{NML}(G_{A})}
            \frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{B})}\nonumber\\
    && =\log 1\cdot\frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{A})}\nonumber\\
    && =\log \frac{P^1_{NML}(B,G'_{B})}{P^1_{NML}(G'_{A})}
             \frac{P^1_{NML}(A,G'_{A})}{P^1_{NML}(G'_{B})}\nonumber\\
 && =s_A^{qNML}(D;G')+s_B^{qNML}(D;G'),\nonumber
\end{eqnarray}
  due to equations $\{A\}\cup G_A = G_B$, $\{B\}\cup G'_B = G'_A$,
  $\{B\}\cup G_B = \{A\} \cup G'_A$, and $G_A = G'_B$ that follow
  easily from the definition of the covered arc.
\end{proof}

@@

Clearly, the qNML score also decomposes by the connected components,
so it remains to show that if the (sub)network $G$ is a tournament,
then for any data $D$, $s^{qNML}(D;G)=\log P_{NML}(D;G)$.  Due to the
score equivalence of both the NML criterion and the qNML criterion, we
may pick such a tournament $G$ that linear ordering defined by $G$
matches the ordering of the data columns, i.e., $i<j$ implies $G_i
\subset G_j$. Now from the definition~(\ref{eqn:qnmldef}) of the qNML
criterion we see that for the tournament $G$, the sum telescopes
leaving us with $s^{qNML}(D;G) = \log P^1_{NML}(D[\cdot,G];G)$, thus
it is enough to show that $P^1_{NML}(D;G)=P_{NML}(D;G)$.
This follows, since for any data vector $x$ in data $D$, the
$P^1(x;\hat\theta(D),G) = P(x;\hat\theta(D),G)$,
where we have denoted by $P^1$ the model that takes $n$-dimensional
vectors to be values of the single (collapsed) categorical variable.
Denoting prefixes of data vector $x$ by $x^{:i}$, and number of times
such a prefix appears in a $N\times n$ datamatrix $D$ by $n_D(x^{:i})$,
($n_D(x^{:0})=N)$, we have
\begin{eqnarray}
  P(D;\hat\theta(D),G) &=& \prod_{j=1}^{N}P(d_j;\hat\theta(D),G) \nonumber \\
  &=& \prod_{j=1}^{N} \prod_{i=1}^{n} \frac{n_D(d_j^{:i})}{n_D(d_j^{:i-1})} \nonumber\\
  &=& \prod_{j=1}^{N} \frac{n_D(d_j^{:n})}{N} \nonumber \\
  &=& P^1(D;\hat\theta^1(D),G).
\end{eqnarray}
Since both $P^1_{NML}$ and $P_{NML}$ are defined in terms of the
maximum likelihood probabilities, the equality above implies the equality
of these distributions.

@@

REWRITE THIS

$dreg(N,q_i,r_i)=reg(N,q_i) -reg(N,r_i q_i)$, where $q$ denotes the number of values in the
collapsed parent variable.  A numerically stable approximation is:
\begin{eqnarray}
\lefteqn{dreg(N,q,r)\approx\frac{1-r^2}{18 N} q^{2}} \\
&&+ \frac{\left(q - 1\right) \sqrt{4 q - 2} + \left(1 - q r\right) \sqrt{4 q r - 2}}{6 \sqrt{N}} \nonumber\\
&&+\frac{q\left(3 N \left(r \log{r}  - \left(r - 1\right) \left(\log{N} + 1\right)\right) + r\right)}{6 N} \nonumber\\
&&+ \frac{1}{12}(\log{(r+\frac{1-r}{q+1})}- 7\log{r}).\nonumber
\end{eqnarray}
